Single-fluorophore orientation determination with multiview polarized illumination: modeling and microscope design

### 2017-11-13 paper/paper/paper.tex
Remove comments and redlines. Submission copy.

\begin{document}
\section*{Reviewer Comments and Author Responses}\hypertarget{test}{}

This document shows the {\color{OE}reviewers' original comments}, our responses,
and the complete paper marked up with the latest revisions. If this document is
viewed as a .pdf, the reader can use the
\hyperlink{test}{{\color{urlblue}links}} in each of our responses to view the
section of the paper that has been changed. After viewing the part of the paper
that was changed, the reader can use the back button ($\cmd + [$ in macOS
Preview.app) to return to the comments and responses.

\section*{Reviewer 1 Comments and Responses}

{\color{OE} Chandler et al. present an exhaustive study of the potential of
  polarization- dependent multiview microscopy for the determination of the
  orientation of individual fluorophores (or any other nano-object with dipolar
  absorption and emission characteristics). Based on earlier work by Fourkas
  (Ref. 15), the authors establish a quantitative description of
  single-fluorophore microscopy that is general enough to cover both
  polarization-dependent excitation and detection, as well non-collinear
  excitation and detection paths and specialized modalities such as light-sheet
  illumination. The authors thus show that an astute choice of the experimental
  configuration allows to determine the 3D orientation of the fluorophore with
  high precision. The latter point is quantified with an information-theoretic
  treatment of the solid-angle uncertainty and illustrated with simulations of
  the resulting performance as a function of several experimental
  parameters. The manuscript is rounded off with a clear discussion of the
  implications for the experimenter, which also takes into account practical
  experimental constraints such as space requirements and the availability of
  high-NA objectives.}

{\color{OE} The manuscript is well-written, thorough and clear, and the
  illustrations are carefully designed to convey the most important conclusions
  in graphical form.  Given the widespread interest in the accurate
  determination of the orientation of single fluorophores, I would expect this
  work to be interesting to many readers, as it carefully addresses both
  theoretical and practical issues related to this topic. I myself have enjoyed
  reading the manuscript and following the reasoning of the authors, and I can
  wholeheartedly recommend publication of this
  nice work in Optics Express.}

{\color{OE} I have a few remarks that the authors may wish to take into account
  to make some
  (very minor) revisions:}

{\color{OE} As far as I know, the correct way to cite the second author of Ref. 17 is D.
A. Vanden Bout. ("Vanden" is part of his last name and should not be
abbreviated.)}\\

\hyperlink{vanden}{\color{urlblue} Corrected.} Thank you for pointing this out.\\

{\color{OE} The rotation matrix of Eq. 6 is identical to the one given by
  Fourkas in Ref.  15. This might surprise readers, because the matrix of
  Fourkas describes the transformation of rays emanating from the focus and
  being refracted by the objective towards the detection polarizer, while Eq. 6
  deals with excitation light being focused, i.e. traveling the reverse optical
  path as far as the objective is concerned. Making the corresponding
  adjustments in Fourkas's treatment, I find that all $\sin\theta$ terms change their
  sign, which affects four matrix elements in Eq. 6. However, using this
  modified matrix results in the exact same of the expressions in Eq. 10, so
  this issue has no practical consequences. It may nevertheless be worthwhile to
  briefly comment on it in order to eliminate a potential source of confusion.}\\

Excellent point. We've changed the signs of the $\sin\theta$ terms in Eq. 6 and \hyperlink{signs}{\color{urlblue} added a sentence} comparing our rotation matrix with Fourkas'.\\

{\color{OE} The expressions for $A$, $B$ and $C$ in Eq. 10 all contain a factor
  of (1-$\cos\alpha$), which can be canceled with the same term in the
  denominator of $D$. I assume the authors have chosen the slightly more
  complicated form of Eq. 10 for didactic purposes (explicitly showing the
  irradiance-conserving factor) and to facilitate comparison with Fourkas's
  results in the subsequent discussion, but it may be worthwhile to mention the
  possibility of further simplification before starting
  numerical calculations.}\\

Very good point. We've \hyperlink{dfactor}{\color{urlblue} added a sentence}
that mentions that these expressions can be simplified.\\

{\color{OE} I think the authors raise an important point when they note that
  Fourkas has incorrectly normalized his expressions for the detected
  intensities. Maybe the authors can briefly comment on whether the underlying
  issue is really only the choice of normalization constant or if this
  discrepancy points to a more fundamental reasoning error that would require an
  angle-dependent correction term.}\\

Fourkas' issue is with incorrect normalization only---there is no
angle-dependent correction required. We've \hyperlink{normalization}{\color{urlblue}
  modified two sentences} to make this clearer.\\

{\color{OE} If one sets $a=b=r$ in Eq. 19 then one does not recover the
  $\gamma$-independent equation of a circle. It seems to me that $x$ and $y$
  have to be exchanged in the
  numerator of the second term of the left-hand side of Eq. 19.}\\

Excellent catch. Thank you. We've \hyperlink{ellipse}{\color{urlblue}
  corrected the equation} and confirmed that when
$a=b=r$ we recover the equation of a circle. We also confirmed that this was a
transcription error only, so it did not affect the ellipse plots throughout the
paper.

\section*{Reviewer 2 Comments and Responses}

{\color{OE} The work by Chandler et al presents an elegant derivation of the
  expressions that characterise the intensity of an single-molecule level
  micrograph containing dipoles with varied orientations, obtained with
  instruments offering multiple views of the same sample. This relates chiefly
  to Light Sheet Microscopes.}

{\color{OE} The derivations in paragraphs 2.2 and 2.3 are scholastic, although
  they provide a correction to a previous work by Fourkas.}

{\color{OE} Figures 4-6 provide the most interesting information to the
  reader, as they describe the theoretical performance of distinct experimental
  arrangements.}

{\color{OE} Surprisingly, no mention is made of the temporal evolution of the
  dipoles. In other words, while these results apply to a fixed condition, it
  would be interesting to discuss a metric concerning the advantages, if any,
  offered by multi view microscopes to capture the temporal dependence of dipole
  orientation evolution.}\\

We thank the reviewer for this valuable comment. We neglected to comment on the
ability of our designs to image dipoles that are rotating during the measurement
time, so we have \hyperlink{temporal}{\color{urlblue} added a paragraph} to
address this omission. We highlight that the modeling in this work only applies
to fixed dipoles, so estimating the orientation of rotating dipoles using these
methods will, in general, lead to incorrect estimates. However, if the dipole
rotates through a small solid angle during the measurement time, the models used
in this paper will give a slightly biased estimate of the average orientation of
the dipole. If the dipoles rotate through a large solid angle during the
measurement time, then further modeling is required to
accurately estimate the dipole orientation.\\

{\color{OE} While this work provides an interesting and novel perspective into
  the problem, what is missing is an experimental validation of the author's
  analytical derivations. This would make the manuscript truly valuable, by
  providing a comparison with experiments where the orientation of the dipoles
  can be fixed, or the rate of change be controlled (i.e. by controlling the
  viscosity). In this respect, an obvious application would be steady state
  anisotropy.}

{\color{OE} We therefore recommend the authors to resubmit the manuscript
  including an experimental section where they test their predictions on at
  least one sample.}\\

We thank the reviewer for the suggestion, but we think that including
experimental results would significantly change the scope of this manuscript. We
wrote this paper to serve as a design guide for those building multiview
polarized light microscopes, and we don't think that experimental results would
help the paper fulfill this role. Also, many of the experimental predictions
made in this paper have already been predicted by Fourkas \cite{fourkas2001} and
verified by others \cite{lu2008}. For these reasons, we have decided to resubmit
the manuscript without experimental results.

\section*{Pre-Production Review Comments and Responses}

{\color{OE} Be sure to include and correct the year on your copyright line to
  the current year.}\\

\hyperlink{copyright}{\color{urlblue} Added.}\\

{\color{OE} For book reference publishers with "Press" as part of publisher
  name, e.g., "Cambridge U. Press," delete "Press" and spell out "University."}\\

\hyperlink{press}{\color{urlblue} Corrected.}\\

{\color{OE} Do not include the city in parentheses with the publisher and year
  for book citations.}\\

\hyperlink{wiley}{\color{urlblue} Changed ``Wiley New York'' to ``Wiley''.}\\

{\color{OE} When you cite a part of a figure in the text, include parentheses
  around the subfigure letter: Fig. 1(b).}\\

Changed throughout and marked in blue.\\

{\color{OE} Do not enclose your figure numbers in parentheses. Parentheses are
  reserved for parts of a figure [Fig. 2(a)] and equations [Eq. (4)].}\\

Changed throughout and marked in blue.\\

{\color{OE} Abbreviate the single word "equation" to "Eq." and the plural word
  "equations" to "Eqs." throughout the text, except for when either word appears
  at the beginning of a sentence.}\\

Changed throughout and marked in blue.\\

{\color{OE} Include parentheses around the equation number in the text:
  Eq. (2), or Eqs. (6)--(8).}\\

Changed throughout.\\

{\color{OE} When citing multiple equations in your text, it should be
  formatted like this: Eqs. (1) and (2) or Eqs. (1)--(3).}\\

Changed throughout.\\

{\color{OE} Include the figure abbreviation each time you cite one in your
  text. Instead of having "(a) and (b)", each time it should be "Fig. 2(a) and
  2(b)".}\\

Changed throughout and marked in blue.\\

{\color{OE} If your paper has funding from the U.S. Department of Energy (DOE)
  or any of its sub-agencies, you are required to deposit a copy of your
  accepted manuscript to the DOE PAGES system. For more information, see
  https://www.osti.gov/elink/am-submission.jsp.}\\

No funding from the DOE.\\

{\color{OE} Please confirm that all of your funding sources have been added
  into your Prism submission in the Fees and Funding step.}\\

Confirmed. \\

{\color{OE} Please make sure to include all of your funding. See the style
  guide for more details at
  https://www.osapublishing.org/submit/templates/default.cfm}\\

Revised the funding section to match the template and mentioned individual funding
sources in the acknowledgments.

\section*{Other Notes}

\hyperlink{typo}{\color{urlblue} Corrected a typo} in the caption of Fig. 1 from $\hat{\mb{e}}_{\text{exc}}$ to
$\hat{\mb{p}}_{\text{exc}}$ for consistency with the rest of the paper.

\pagebreak

\title{Single-fluorophore orientation determination with multiview polarized
  illumination: modeling and microscope design}

@@

%%%%%%%%%%%%%%%%%%% abstract and OCIS codes %%%%%%%%%%%%%%%%
\begin{abstract}
  We investigate the use of polarized illumination in multiview microscopes for
  determining the orientation of single-molecule fluorescence transition
  dipoles. First, we relate the orientation of single dipoles to measurable
  intensities in multiview microscopes and develop an information-theoretic
  metric---the solid-angle uncertainty---to compare the ability of multiview
  microscopes to estimate the orientation of single dipoles. Next, we compare
  a broad class of microscopes using this metric---single- and dual-view
  microscopes with varying illumination polarization, illumination numerical
  aperture (NA), detection NA, obliquity, asymmetry, and exposure. We find that
  multi-view microscopes can measure all dipole orientations, while the
  orientations measurable with single-view microscopes is halved because of
  symmetries in the detection process. We also find that choosing a small
  illumination NA and a large detection NA are good design choices, that
  multiview microscopes can benefit from oblique illumination and detection, and
  that asymmetric NA microscopes can benefit from exposure asymmetry.
\end{abstract}
\hypertarget{copyright}{}

@@

\bibitem{lu2008}
C.-Y. Lu and D.~A.~\hypertarget{vanden}{{\color{urlblue} Vanden}} Bout, \enquote{Analysis of orientational dynamics of
  single fluorophore trajectories from three-angle polarization experiments,}
  J. Chem. Phys. \textbf{128}, 244501 (2008).

@@

\bibitem{anderson1958}
  T.~W. Anderson, \emph{An introduction to multivariate statistical analysis}
(\hypertarget{wiley}{{\color{urlblue}
Wiley}}, 1958).

@@

\bibitem{nov2006}
L.~Novotny and B.~Hecht, \emph{Principles of Nano-Optics} (\hypertarget{press}
  {{\color{urlblue}
Cambridge University}}, 2006).

@@

We consider a single molecule with a fixed absorption dipole moment
$\hat{\bs{\mu}}_{\text{abs}}$ that we express in spherical coordinates $\Theta$
and $\Phi$ {\color{urlblue}[}see {\color{urlblue}Fig.} \ref{fig:coordinates}{\color{urlblue}(a)}{\color{urlblue}]} as
\begin{align}
  \hat{\bs{\mu}}_{\text{abs}}(\Theta, \Phi) = \sin\Theta\cos\Phi\hat{\mb{x}} + \sin\Theta\sin\Phi\hat{\mb{y}} + \cos\Theta\hat{\mb{z}} \label{eq:mu}.
\end{align}
We place the molecule in the focal plane of an ideal, aplanatic,
polarization-preserving objective lens with its optical axis aligned with the
$\mh{z}$ axis. Next, we place a polarizer in the aperture plane with a variable
transmission axis $\hat{\mb{p}}_{\text{exc}}$ that we express as
\begin{align}
 \hat{\mb{p}}_{\text{exc}}(\phi_{\text{exc}}) = \cos\phi_{\text{exc}}\hat{\mb{x}} +
\sin\phi_{\text{exc}}\hat{\mb{y}},\label{eq:excfield}
\end{align}
where $\phi_{\text{exc}}$ is the angle between the transmission axis of the
polarizer and the positive $\mh{x}$ axis {\color{urlblue}[}see {\color{urlblue}Fig.}
\ref{fig:coordinates}{\color{urlblue}(b)}{\color{urlblue}]}. Finally, we place a spatially incoherent and spatially
uniform light source (or its image) in the aperture plane to illuminate the
focal plane. In this geometry each point in the sample is illuminated by a range
of angles that can be changed by adjusting the radius of the diaphragm in the
aperture plane. We denote the set of illuminated angles by $\Omega$ {\color{urlblue}[}see {\color{urlblue}Fig.}
\ref{fig:coordinates}{\color{urlblue}(d)}{\color{urlblue}]}. If we use unit vectors $\hat{\mb{r}}$ expressed in
spherical coordinates $\theta$ and $\phi$ {\color{urlblue}[}see {\color{urlblue}Fig.} \ref{fig:coordinates}{\color{urlblue}(c)}{\color{urlblue}]}
as
  \begin{align}
   \hat{\mb{r}}(\theta, \phi) = \sin\theta\cos\phi\hat{\mb{x}} + \sin\theta\sin\phi\hat{\mb{y}} + \cos\theta\hat{\mb{z}}, 
  \end{align}
  then the set of angles $\Omega$ can be expressed as
  \begin{align}
  \Omega = \{\phi, \theta\ |\ 0 < \phi \leq 2\pi, 0 < \theta \leq \alpha\},
  \end{align}
  where $\alpha$ is the angle between the optical axis and the most oblique
  illuminating ray. Equivalently, $\alpha$ can be expressed in terms of NA using
  \begin{align}
    \text{NA} = n\sin\alpha, 
  \end{align}
  where $n$ is the index of refraction of the sample medium. Note that the NA of
  the objective is always greater than or equal to the illumination NA because
  we can underfill the aperture plane.

  The objective focuses the light by applying a position-dependent rotation to
  the electric field which we model by multiplying the incident electric field
  $\hat{\mb{p}}_{\text{exc}}$ by a position-dependent rotation matrix
  \begin{align}
  \mb{R}(\mh{r}) &= \begin{bmatrix} \cos\theta\cos^2\phi + \sin^2\phi & (\cos\theta -1)\sin\phi\cos\phi & \sin\theta\cos\phi\\ (\cos\theta - 1)\sin\phi\cos\phi & \cos\theta\sin^2\phi + \cos^2\phi & \sin\theta\sin\phi \\ -\sin\theta\cos\phi& -\sin\theta\sin\phi & \cos\theta \end{bmatrix} \label{eq:rot_mat}.
  \end{align}
  \hypertarget{signs}
  {{\color{urlblue}Note the difference of signs between Eq.
  (\ref{eq:rot_mat}) and the rotation matrix given by
    Fourkas---our matrix rotates rays propagating towards the sample while
    Fourkas' matrix rotates rays propagating away from the sample.}}

@@

We substitute {\color{urlblue}Eqs.} (\ref{eq:mu})--(\ref{eq:rot_mat}) into
{\color{urlblue}Eq.} (\ref{eq:abs}), evaluate the integrals, and simplify to
express the absorption efficiency in scalar notation as
\begin{align}
  \eta_{\text{abs}} &= D\{A + B\sin^{2}{\Theta} + C\sin^{2}{\Theta} \cos{[2 (\Phi - \phi_{\text{exc}}})]\}\label{eq:scalarabs}
\end{align}
where
\begin{subequations}
\begin{align}
  A &= \frac{1}{4} - \frac{3}{8} \cos{\alpha } + \frac{1}{8} \cos^{3}{\alpha }\\
  B &= \frac{3}{16} \cos{\alpha } - \frac{3}{16} \cos^{3}{\alpha }\\
  C &= \frac{7}{32} - \frac{3}{32} \cos{\alpha } - \frac{3}{32} \cos^{2}{\alpha } - \frac{1}{32} \cos^{3}{\alpha}\\
  D &= \frac{4}{3(1 - \cos\alpha)}.
\end{align}\label{eq:coefficients}%
\end{subequations}
Note that the factor $D$ keeps the specimen irradiance constant for any value of
$\alpha$. \hypertarget{dfactor}{{\color{urlblue} We have kept $D$ as a
    multiplicative factor in Eq. (\ref{eq:scalarabs}) to facilitate comparison
    with Fourkas' results, but Eq. (\ref{eq:scalarabs}) can be simplified
    further by distributing $D$.}}

We can extend these results to light-sheet illumination created by scanning a
laser. If we shine a narrow, collimated, laser beam onto a galvanometer in the
aperture plane and scan the galvanometer, we create a weakly focused light sheet
in the sample. If we scan the beam in the sample with a velocity much less than
the beam waist radius divided by the coherence time, then the coherence of the
laser is rendered ineffectual by the scanning---the light sheet is effectively
composed of many incoherent beams. If we also ensure that the laser beam is
weakly focused, we can ignore the excitation caused by the small fields aligned
with the optical axis. Under slowly-scanned and weakly-focused light-sheet
illumination, dipoles in the sample will be excited as if a single plane wave
was incident. Therefore, we can find the absorption efficiency of a dipole in
this case by taking the limit of {\color{urlblue}Eq.} (\ref{eq:scalarabs}) as
$\alpha \rightarrow 0$ {\color{urlblue}[}or by plugging {\color{urlblue}Eqs.}
(\ref{eq:mu}) and (\ref{eq:excfield}) into {\color{urlblue}Eq.}
(\ref{eq:singleray}){\color{urlblue}]} giving
\begin{align}
  \eta_{\text{abs}}^{\text{ray}} &= \sin^2\Theta\cos^2(\Phi - \phi_{\text{exc}}),
\end{align}
which is recognizable as Malus' law generalized to three dimensions. This means
that light-sheet illumination is approximately equivalent to low-NA illumination
with respect to the dipole orientations they excite. Notice that the
absorption efficiency can take its maximum range of values
($\eta_{\text{abs}} \in [0, 1]$) when $\alpha=0$.

@@

To calculate the detection efficiency without a polarizer, we use {\color{urlblue}Eq.}
(\ref{eq:det_pol}) and take the sum of polarized detection efficiencies with
orthogonal polarizer orientations to find
\begin{align}
  \eta_{\text{det}} = 2(A + B\sin^2\Theta). \label{eq:scalardet}
\end{align}
In this case the detection efficiency only depends on $\Theta$, not $\Phi$,
because there is no detection polarizer. We have assumed that the absorption
dipole moment is equal to the emission dipole moment.

Note that $A$, $B$, and $C$ in {\color{urlblue}Eq.} (\ref{eq:coefficients}) are a
factor of $\frac{3}{2}$ larger than the expressions given by Fourkas. We found
that Fourkas' expressions were incorrectly normalized (the limit of
$\eta_{\text{det}}$ as $\alpha\rightarrow \frac{\pi}{2}$ should be
$\frac{1}{2}$), and the extra factor of $\frac{3}{2}$ corrects the
error. \hypertarget{normalization}{{\color{urlblue} The extra factor of
    $\frac{3}{2}$ is a constant multiplicative factor, so the}} correction
{\color{urlblue} is not angle-dependent which means that it} does not affect
Fourkas' orientation reconstruction. {\color{urlblue}However, the correction}
does mean that Fourkas' algorithm underpredicts the total emitted intensity by a
factor of $\frac{3}{2}$.

@@

\begin{figure}[H]
\centering\includegraphics[width=\textwidth]{single-frame}
\caption{Representative examples of single intensity measurements. Black dots
  indicate where the Cartesian unit vectors intersect the unit sphere. \newline \newline \textbf{Columns left to right:} 1) schematics where
  the solid line encloses the illumination solid angle, the dashed line encloses
  the detection solid angle, and the arrow indicates the transmission axis of
  the illumination polarizer; 2) the absorption efficiency, see {\color{urlblue}Eq.}
  (\ref{eq:scalarabs}); 3) the detection efficiency, see {\color{urlblue}Eq.}
  (\ref{eq:scalardet}); 4) the total efficiency, the product of the absorption
  and detection efficiencies.\newline \newline \textbf{Rows top to bottom:} 1)
  coincident illumination ($\text{NA} = 1.1$ with $x$-polarized light) and
  detection (NA = 1.1); 2) non-coincident orthogonal illumination
  ($\text{NA} = 0.8$) and detection ($\text{NA} = 0.8$); 3) non-coincident
  135${}^{\circ}$-separated illumination ($\text{NA} = 0.8$) and detection
  ($\text{NA} = 0.8$). All simulations use $n=1.33$.}
  \label{fig:single-frame}
\end{figure}

@@

We also use the Fisher information matrix to calculate uncertainty ellipses for
each microscope and dipole orientation. The general equation of an ellipse
with axis radii $a$ and $b$ and rotation angle $\gamma$ is
\hypertarget{ellipse}{{\color{urlblue} 
\begin{align}
  \frac{(x\cos\gamma + y\sin\gamma)^2}{a^2} + \frac{(x\sin\gamma - y\cos\gamma)^2}{b^2} =1. 
\end{align}}}%
We calculate the ellipse parameters in terms of the Fisher information matrix
using the following set of equations
\begin{align}
  \sigma_x &= \sqrt{\mb{F}^{-1}_{0,0}} \label{eq:first}\\
  \sigma_y &= \sin\Theta\sqrt{\mb{F}^{-1}_{1,1}} \label{eq:second}\\ 
  \sigma_{xy} &= \sin\Theta\sqrt{\mb{F}^{-1}_{1,0}} \label{eq:third}\\  
  a^2 &= \frac{\sigma_x^2 + \sigma_y^2}{2} + \sqrt{\frac{(\sigma_x^2 - \sigma_y^2)^2}{4} + \sigma_{xy}^2}\\
  b^2 &= \frac{\sigma_x^2 + \sigma_y^2}{2} - \sqrt{\frac{(\sigma_x^2 - \sigma_y^2)^2}{4} + \sigma_{xy}^2}\\
  \gamma &= \frac{1}{2}\text{arctan}\left(\frac{2\sigma_{xy}}{\sigma_x^2 - \sigma_y^2}\right). \label{eq:last}
\end{align}
Equations (\ref{eq:first})--(\ref{eq:last}) are identical to those found in
\cite{coe2009} with an additional factor of $\sin\Theta$ on {\color{urlblue}Eqs.}
(\ref{eq:second}) and (\ref{eq:third}) to account for the non-Euclidean geometry of
the sphere. Instead of plotting ellipses with widely varying radii on the same
figure, we chose a constant major axis radius. This means that the uncertainty
ellipses indicate the relative size of uncertainty in different directions, not
the absolute size of the uncertainty.

>>>>

\begin{document}
\title{Single-fluorophore orientation determination with multiview polarized
  illumination: modeling and microscope design}

@@

%%%%%%%%%%%%%%%%%%% abstract and OCIS codes %%%%%%%%%%%%%%%%
\begin{abstract}
  We investigate the use of polarized illumination in multiview microscopes for
  determining the orientation of single-molecule fluorescence transition
  dipoles. First, we relate the orientation of single dipoles to measurable
  intensities in multiview microscopes and develop an information-theoretic
  metric---the solid-angle uncertainty---to compare the ability of multiview
  microscopes to estimate the orientation of single dipoles. Next, we compare
  a broad class of microscopes using this metric---single- and dual-view
  microscopes with varying illumination polarization, illumination numerical
  aperture (NA), detection NA, obliquity, asymmetry, and exposure. We find that
  multi-view microscopes can measure all dipole orientations, while the
  orientations measurable with single-view microscopes is halved because of
  symmetries in the detection process. We also find that choosing a small
  illumination NA and a large detection NA are good design choices, that
  multiview microscopes can benefit from oblique illumination and detection, and
  that asymmetric NA microscopes can benefit from exposure asymmetry.
\end{abstract}

@@

\bibitem{lu2008}
C.-Y. Lu and D.~A.~Vanden Bout, \enquote{Analysis of orientational dynamics of
  single fluorophore trajectories from three-angle polarization experiments,}
  J. Chem. Phys. \textbf{128}, 244501 (2008).

@@

\bibitem{anderson1958}
  T.~W. Anderson, \emph{An introduction to multivariate statistical analysis}
(Wiley, 1958).

@@

\bibitem{nov2006}
L.~Novotny and B.~Hecht, \emph{Principles of Nano-Optics} (Cambridge University, 2006).

@@

We consider a single molecule with a fixed absorption dipole moment
$\hat{\bs{\mu}}_{\text{abs}}$ that we express in spherical coordinates $\Theta$
and $\Phi$ [see Fig. \ref{fig:coordinates}(a)] as
\begin{align}
  \hat{\bs{\mu}}_{\text{abs}}(\Theta, \Phi) = \sin\Theta\cos\Phi\hat{\mb{x}} + \sin\Theta\sin\Phi\hat{\mb{y}} + \cos\Theta\hat{\mb{z}} \label{eq:mu}.
\end{align}
We place the molecule in the focal plane of an ideal, aplanatic,
polarization-preserving objective lens with its optical axis aligned with the
$\mh{z}$ axis. Next, we place a polarizer in the aperture plane with a variable
transmission axis $\hat{\mb{p}}_{\text{exc}}$ that we express as
\begin{align}
 \hat{\mb{p}}_{\text{exc}}(\phi_{\text{exc}}) = \cos\phi_{\text{exc}}\hat{\mb{x}} +
\sin\phi_{\text{exc}}\hat{\mb{y}},\label{eq:excfield}
\end{align}
where $\phi_{\text{exc}}$ is the angle between the transmission axis of the
polarizer and the positive $\mh{x}$ axis [see Fig.
\ref{fig:coordinates}(b)]. Finally, we place a spatially incoherent and
spatially uniform light source (or its image) in the aperture plane to
illuminate the focal plane. In this geometry each point in the sample is
illuminated by a range of angles that can be changed by adjusting the radius of
the diaphragm in the aperture plane. We denote the set of illuminated angles by
$\Omega$ [see Fig. \ref{fig:coordinates}(d)]. If we use unit vectors
$\hat{\mb{r}}$ expressed in spherical coordinates $\theta$ and $\phi$ [see
Fig. \ref{fig:coordinates}{(c)] as
  \begin{align}
   \hat{\mb{r}}(\theta, \phi) = \sin\theta\cos\phi\hat{\mb{x}} + \sin\theta\sin\phi\hat{\mb{y}} + \cos\theta\hat{\mb{z}}, 
  \end{align}
  then the set of angles $\Omega$ can be expressed as
  \begin{align}
  \Omega = \{\phi, \theta\ |\ 0 < \phi \leq 2\pi, 0 < \theta \leq \alpha\},
  \end{align}
  where $\alpha$ is the angle between the optical axis and the most oblique
  illuminating ray. Equivalently, $\alpha$ can be expressed in terms of NA using
  \begin{align}
    \text{NA} = n\sin\alpha, 
  \end{align}
  where $n$ is the index of refraction of the sample medium. Note that the NA of
  the objective is always greater than or equal to the illumination NA because
  we can underfill the aperture plane.

  The objective focuses the light by applying a position-dependent rotation to
  the electric field which we model by multiplying the incident electric field
  $\hat{\mb{p}}_{\text{exc}}$ by a position-dependent rotation matrix
  \begin{align}
  \mb{R}(\mh{r}) &= \begin{bmatrix} \cos\theta\cos^2\phi + \sin^2\phi & (\cos\theta -1)\sin\phi\cos\phi & \sin\theta\cos\phi\\ (\cos\theta - 1)\sin\phi\cos\phi & \cos\theta\sin^2\phi + \cos^2\phi & \sin\theta\sin\phi \\ -\sin\theta\cos\phi& -\sin\theta\sin\phi & \cos\theta \end{bmatrix} \label{eq:rot_mat}.
  \end{align}
  Note the difference of signs between Eq.
  (\ref{eq:rot_mat}) and the rotation matrix given by
    Fourkas---our matrix rotates rays propagating towards the sample while
    Fourkas' matrix rotates rays propagating away from the sample.

@@

We substitute Eqs. (\ref{eq:mu})--(\ref{eq:rot_mat}) into
Eq. (\ref{eq:abs}), evaluate the integrals, and simplify to
express the absorption efficiency in scalar notation as
\begin{align}
  \eta_{\text{abs}} &= D\{A + B\sin^{2}{\Theta} + C\sin^{2}{\Theta} \cos{[2 (\Phi - \phi_{\text{exc}}})]\}\label{eq:scalarabs}
\end{align}
where
\begin{subequations}
\begin{align}
  A &= \frac{1}{4} - \frac{3}{8} \cos{\alpha } + \frac{1}{8} \cos^{3}{\alpha }\\
  B &= \frac{3}{16} \cos{\alpha } - \frac{3}{16} \cos^{3}{\alpha }\\
  C &= \frac{7}{32} - \frac{3}{32} \cos{\alpha } - \frac{3}{32} \cos^{2}{\alpha } - \frac{1}{32} \cos^{3}{\alpha}\\
  D &= \frac{4}{3(1 - \cos\alpha)}.
\end{align}\label{eq:coefficients}%
\end{subequations}
Note that the factor $D$ keeps the specimen irradiance constant for any value of
$\alpha$. We have kept $D$ as a multiplicative factor in
Eq. (\ref{eq:scalarabs}) to facilitate comparison with Fourkas' results, but
Eq. (\ref{eq:scalarabs}) can be simplified further by distributing $D$.

We can extend these results to light-sheet illumination created by scanning a
laser. If we shine a narrow, collimated, laser beam onto a galvanometer in the
aperture plane and scan the galvanometer, we create a weakly focused light sheet
in the sample. If we scan the beam in the sample with a velocity much less than
the beam waist radius divided by the coherence time, then the coherence of the
laser is rendered ineffectual by the scanning---the light sheet is effectively
composed of many incoherent beams. If we also ensure that the laser beam is
weakly focused, we can ignore the excitation caused by the small fields aligned
with the optical axis. Under slowly-scanned and weakly-focused light-sheet
illumination, dipoles in the sample will be excited as if a single plane wave
was incident. Therefore, we can find the absorption efficiency of a dipole in
this case by taking the limit of Eq. (\ref{eq:scalarabs}) as
$\alpha \rightarrow 0$ [or by plugging Eqs. (\ref{eq:mu}) and
(\ref{eq:excfield}) into Eq. (\ref{eq:singleray})] giving
\begin{align}
  \eta_{\text{abs}}^{\text{ray}} &= \sin^2\Theta\cos^2(\Phi - \phi_{\text{exc}}),
\end{align}
which is recognizable as Malus' law generalized to three dimensions. This means
that light-sheet illumination is approximately equivalent to low-NA illumination
with respect to the dipole orientations they excite. Notice that the
absorption efficiency can take its maximum range of values
($\eta_{\text{abs}} \in [0, 1]$) when $\alpha=0$.

@@

To calculate the detection efficiency without a polarizer, we use Eq.
(\ref{eq:det_pol}) and take the sum of polarized detection efficiencies with
orthogonal polarizer orientations to find
\begin{align}
  \eta_{\text{det}} = 2(A + B\sin^2\Theta). \label{eq:scalardet}
\end{align}
In this case the detection efficiency only depends on $\Theta$, not $\Phi$,
because there is no detection polarizer. We have assumed that the absorption
dipole moment is equal to the emission dipole moment.

Note that $A$, $B$, and $C$ in Eq. (\ref{eq:coefficients}) are a factor of
$\frac{3}{2}$ larger than the expressions given by Fourkas. We found that
Fourkas' expressions were incorrectly normalized (the limit of
$\eta_{\text{det}}$ as $\alpha\rightarrow \frac{\pi}{2}$ should be
$\frac{1}{2}$), and the extra factor of $\frac{3}{2}$ corrects the error. The
extra factor of $\frac{3}{2}$ is a constant multiplicative factor, so the
correction is not angle-dependent which means that it does not affect Fourkas'
orientation reconstruction. However, the correction does mean that Fourkas'
algorithm underpredicts the total emitted intensity by a factor of
$\frac{3}{2}$.

@@

\begin{figure}[H]
\centering\includegraphics[width=\textwidth]{single-frame}
\caption{Representative examples of single intensity measurements. Black dots
  indicate where the Cartesian unit vectors intersect the unit sphere. \newline
  \newline \textbf{Columns left to right:} 1) schematics where the solid line
  encloses the illumination solid angle, the dashed line encloses the detection
  solid angle, and the arrow indicates the transmission axis of the illumination
  polarizer; 2) the absorption efficiency, see Eq.  (\ref{eq:scalarabs}); 3) the
  detection efficiency, see Eq.  (\ref{eq:scalardet}); 4) the total efficiency,
  the product of the absorption and detection efficiencies.\newline \newline
  \textbf{Rows top to bottom:} 1) coincident illumination ($\text{NA} = 1.1$
  with $x$-polarized light) and detection (NA = 1.1); 2) non-coincident
  orthogonal illumination ($\text{NA} = 0.8$) and detection ($\text{NA} = 0.8$);
  3) non-coincident 135${}^{\circ}$-separated illumination ($\text{NA} = 0.8$)
  and detection ($\text{NA} = 0.8$). All simulations use $n=1.33$.}
  \label{fig:single-frame}
\end{figure}

@@

We also use the Fisher information matrix to calculate uncertainty ellipses for
each microscope and dipole orientation. The general equation of an ellipse
with axis radii $a$ and $b$ and rotation angle $\gamma$ is
\begin{align}
  \frac{(x\cos\gamma + y\sin\gamma)^2}{a^2} + \frac{(x\sin\gamma - y\cos\gamma)^2}{b^2} =1. 
\end{align}%
We calculate the ellipse parameters in terms of the Fisher information matrix
using the following set of equations
\begin{align}
  \sigma_x &= \sqrt{\mb{F}^{-1}_{0,0}} \label{eq:first}\\
  \sigma_y &= \sin\Theta\sqrt{\mb{F}^{-1}_{1,1}} \label{eq:second}\\ 
  \sigma_{xy} &= \sin\Theta\sqrt{\mb{F}^{-1}_{1,0}} \label{eq:third}\\  
  a^2 &= \frac{\sigma_x^2 + \sigma_y^2}{2} + \sqrt{\frac{(\sigma_x^2 - \sigma_y^2)^2}{4} + \sigma_{xy}^2}\\
  b^2 &= \frac{\sigma_x^2 + \sigma_y^2}{2} - \sqrt{\frac{(\sigma_x^2 - \sigma_y^2)^2}{4} + \sigma_{xy}^2}\\
  \gamma &= \frac{1}{2}\text{arctan}\left(\frac{2\sigma_{xy}}{\sigma_x^2 - \sigma_y^2}\right). \label{eq:last}
\end{align}
Equations (\ref{eq:first})--(\ref{eq:last}) are identical to those found in
\cite{coe2009} with an additional factor of $\sin\Theta$ on Eqs.
(\ref{eq:second}) and (\ref{eq:third}) to account for the non-Euclidean geometry of
the sphere. Instead of plotting ellipses with widely varying radii on the same
figure, we chose a constant major axis radius. This means that the uncertainty
ellipses indicate the relative size of uncertainty in different directions, not
the absolute size of the uncertainty.
